---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---

```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(scales)
Sys.setenv(TZ='UTC')
```

## Loading and preprocessing the data

The step data is loaded from a '[activity.csv](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)' file provided by the instructor. The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

A fourth column was appended using the code below to represent a POSIXct datetiem value, where the date and five-minute time of day are pasted together and parsed using the `strptime()` function.


```{r loadActivityData}

activityData <- read.csv(
    file = 'activity.csv',
    colClasses = c('integer', 'character','character'),
    header = TRUE,
    na.strings = c('NA')
)

## convert the date string to a POSIX value
activityData$date <- as.POSIXct(
    strptime(activityData$date, "%Y-%m-%d"),
    tz="UTC"
)

## convert the interval string to a POSIX value
activityData$interval <- as.POSIXct(
    strptime(
        paste(
                as.Date(as.POSIXct(0,
                                   origin = "1970-01-01",
                                   tz = "UTC")),
                " ",
                substr(sprintf("%04s", activityData$interval), 1, 2),
                ":",
                substr(sprintf("%04s", activityData$interval), 3, 4),
                ":00",
                sep = ""
        ),
        "%Y-%m-%d %H:%M:%S"
    ),
    tz = "UTC"
)

## combine the date and time into a single value
activityData$dateTime <- as.POSIXct(
    strptime(
        paste(  activityData$date,
                strftime(activityData$interval, format = "%H:%M:%S", tz = "UTC"),
                sep = " "
        ),
        "%Y-%m-%d %H:%M:%S"
    ),
    tz = "UTC"
)

```


## What is mean total number of steps taken per day?

```{r totalStepsPerDay}

totalSteps <- aggregate(
    formula = steps ~ date,
    data = activityData,
    FUN = sum,
    na.action = na.omit
)


dailyStepsAvg <- mean(totalSteps$steps)
dailyStepsMedian <- median(totalSteps$steps)
```
```{r plotTotalStepsPerDay}

plotTotalSteps <- ggplot(totalSteps, aes(date))
plotTotalSteps + geom_bar(
    aes(weight = steps),
    color = "blue",
    fill = "lightblue"
    ) +
    labs(title = "Total Steps Per Day",
        subtitle = "No Imputation",
        x = "Date",
        y = "Number of Steps"
    )

```
The average number of steps per day is **`r sprintf("%.1f", dailyStepsAvg)`**.

The median number of steps per day is **`r sprintf("%i", dailyStepsMedian)`**.


## What is the average daily activity pattern?

I guess I need to make a time-series plot of the inter-average for multiple days here.

```{r computeIntervalAverages}

# create interval averages for all days
intervalStepsAvg <- aggregate(
    formula = steps ~ interval,
    data = activityData,
    FUN = mean,
    na.action = na.omit
)

# determine which interval has the largest number of steps
intervalStepsMax <- intervalStepsAvg[which.max(intervalStepsAvg$steps),]

```
```{r plotIntervalAverages}

# plot 'em intervals up real good
plotIntervalSteps <- ggplot(intervalStepsAvg, aes(interval, steps))
plotIntervalSteps +
    geom_path(color = "blue") +
    labs(title = "Daily Average Steps",
         subtitle = "No Imputation",
         x = "Time of Day",
         y = "Average Steps") +
    scale_x_datetime(labels = date_format("%H:%M", tz = "UTC"))

```

The 5-minute interval containing the most steps is **`r sub("^\\s+", "", strftime(intervalStepsMax$interval, format = "%l:%M %p", tz = "UTC"))`** with **`r sprintf("%.0f", intervalStepsMax$steps)`** steps.

## Imputing missing values

What I want to do here is append an additional column to the data frame, where the original `steps` values is used if not `NA`, and if the value is missing, substitute the average five-minute value for the entire data set that is appropriate for that specific interval.

The total number of `NA` intervals is **`r nrow(activityData[is.na(activityData$steps),])`**.

```{r imputeOhNooo}

# append an imputed number of steps to the original activity data.
activityData$stepsImpute <- cbind(
    apply(activityData[,c("steps","interval")],
        1,
        function(x) as.integer(
            ifelse(
                is.na(x['steps']),
                    round(
                        intervalStepsAvg[
                            which(intervalStepsAvg$interval == x['interval']),
                        ]$steps,
                        digits = 0),
                x['steps']
            )
        )
    )
)

```
```{r aggregateStepsImpute}

# aggregate the imputed total steps.
totalStepsImpute <- aggregate(
    formula = stepsImpute ~ date,
    data = activityData,
    FUN = sum
)

names(totalStepsImpute) <- c("date","stepsImpute")

dailyStepsImputeAvg <- mean(totalStepsImpute$stepsImpute)
dailyStepsImputeMedian <- median(totalStepsImpute$stepsImpute)

```
```{r plotStepsImpute}

# make a graph, recieve cookie
plotTotalStepsImpute <- ggplot(totalStepsImpute, aes(date))
plotTotalStepsImpute + geom_bar(
    aes(weight = stepsImpute),
    color = "firebrick",
    fill = "firebrick1"
    ) +
    labs(title = "Total Steps Per Day",
        subtitle = "Data Imputed from Interval Average",
        x = "Date",
        y = "Number of Steps"
    )

```

After imputation, the average number of steps per day is **`r sprintf("%.1f", dailyStepsImputeAvg)`**.

The median number of steps per day is **`r sprintf("%i", dailyStepsImputeMedian)`**.

## Are there differences in activity patterns between weekdays and weekends?

Yes, I am wasting my stupid weekend working on this project, while instead, I get paid to work on weekdays.

```{r processWeekendIdiocy}

#add a day of week, why not?
activityData$dayOfWeek <- cbind(weekdays(activityData$dateTime))

#then classify that day from above as a weekday/weekend in another column
activityData$weekend <- cbind(
    apply(
        activityData[,"dayOfWeek"],
        1,
        function(x) ifelse(x %in% c("Saturday","Sunday"), "Weekend", "Weekday")
    )
)

# calculate interval averages of imputed steps for weekdays, yay!
intervalStepsAvgWeekday <- aggregate(
    formula = stepsImpute ~ interval,
    data = activityData[which(activityData$weekend == "Weekday"),],
    FUN = mean,
    na.action = na.omit
)

names(intervalStepsAvgWeekday) <- c("interval","stepsImpute")

# I should probably have defined a function to aggregate this!
intervalStepsAvgWeekend <- aggregate(
    formula = stepsImpute ~ interval,
    data = activityData[which(activityData$weekend == "Weekend"),],
    FUN = mean,
    na.action = na.omit
)

names(intervalStepsAvgWeekend) <- c("interval","stepsImpute")

```
```{r multiplot}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```
```{r plotWeekendIdiocy}

# make another graph, probably should have written a function for this, too!
plotIntervalStepsWeekday <- ggplot(intervalStepsAvgWeekday, aes(interval, stepsImpute))
plotAvgWeekday <- plotIntervalStepsWeekday +
    geom_path(color = "blue") +
    labs(title = "Daily Average Steps",
         subtitle = "Weekday",
         x = "Time of Day",
         y = "Average Steps") +
    scale_x_datetime(labels = date_format("%H:%M", tz = "UTC"))

# yeah, I definitely should have written a function to do this!
plotIntervalStepsWeekend <- ggplot(intervalStepsAvgWeekend, aes(interval, stepsImpute))
plotAvgWeekend <- plotIntervalStepsWeekend +
    geom_path(color = "blue") +
    labs(title = "Daily Average Steps",
         subtitle = "Weekend",
         x = "Time of Day",
         y = "Average Steps") +
    scale_x_datetime(labels = date_format("%H:%M", tz = "UTC"))

multiplot(plotAvgWeekday, plotAvgWeekend, cols = 1)

```