---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---

```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(scales)
```

## Loading and preprocessing the data

The step data is loaded from a '[activity.csv](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)' file provided by the instructor. The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

A fourth column was appended using the code below to represent a POSIXct datetiem value, where the date and five-minute time of day are pasted together and parsed using the `strptime()` function.


```{r loadActivityData}

activityData <- read.csv(
    file = 'activity.csv',
    colClasses = c('integer', 'character','character'),
    header = TRUE,
    na.strings = c('NA')
)

## convert the date string to a POSIX value
activityData$date <- as.POSIXct(
    strptime(activityData$date, "%Y-%m-%d"),
    tz="UTC"
)

## convert the interval string to a POSIX value
activityData$interval <- as.POSIXct(
    strptime(
        paste(
                as.Date(as.POSIXct(0,
                                   origin = "1970-01-01",
                                   tz = "UTC")),
                " ",
                substr(sprintf("%04s", activityData$interval), 1, 2),
                ":",
                substr(sprintf("%04s", activityData$interval), 3, 4),
                ":00",
                sep = ""
        ),
        "%Y-%m-%d %H:%M:%S"
    ),
    tz = "UTC"
)

## combine the date and time into a single value
activityData$dateTime <- as.POSIXct(
    strptime(
        paste(  activityData$date,
                strftime(activityData$interval, format = "%H:%M:%S", tz = "UTC"),
                sep = " "
        ),
        "%Y-%m-%d %H:%M:%S"
    ),
    tz = "UTC"
)

```


## What is mean total number of steps taken per day?

```{r totalStepsPerDay}

totalSteps <- aggregate(
    formula = steps ~ date,
    data = activityData,
    FUN = sum,
    na.action = na.omit
)


dailyStepsAvg <- mean(totalSteps$steps)
dailyStepsMedian <- median(totalSteps$steps)


plotTotalSteps <- ggplot(totalSteps, aes(date))
plotTotalSteps + geom_bar(
    aes(weight = steps),
    color = "blue",
    fill = "lightblue"
    ) +
    labs(title = "Total Steps Per Day",
        subtitle = "No Imputation",
        x = "Date",
        y = "Number of Steps"
    )

```
The average number of steps per day is **`r sprintf("%.1f", dailyStepsAvg)`**.

The median number of steps per day is **`r sprintf("%i", dailyStepsMedian)`**.


## What is the average daily activity pattern?

I guess I need to make a time-series plot of the inter-average for multiple days here.

```{r dailyAverage}

# create interval averages for all days
intervalStepsAvg <- aggregate(
    formula = steps ~ interval,
    data = activityData,
    FUN = mean,
    na.action = na.omit
)

# determine which interval has the largest number of steps
intervalStepsMax <- intervalStepsAvg[which.max(intervalStepsAvg$steps),]

# plot 'em intervals up real good
plotIntervalSteps <- ggplot(intervalStepsAvg, aes(interval, steps))
plotIntervalSteps +
    geom_path(color = "blue") +
    labs(title = "Daily Average Steps",
         subtitle = "No Imputation",
         x = "Time of Day",
         y = "Average Steps") +
    scale_x_datetime(labels = date_format("%H:%M", tz = "UTC"))

```

The 5-minute interval containing the most steps is **`r sub("^\\s+", "", strftime(intervalStepsMax$interval, format = "%l:%M %p", tz = "UTC"))`** with **`r sprintf("%.0f", intervalStepsMax$steps)`** steps.

## Imputing missing values

What I want to do here is append an additional column to the data frame, where the original `steps` values is used if not `NA`, and if the value is missing, substitute the median five-minute value for the entire data set that is appropriate for that specific interval.

## Are there differences in activity patterns between weekdays and weekends?
